Topological Equivariant Artist Model

### 2021-03-26 slides/main.tex
lets not delete my appendix

\begin{frame}
	\titlepage
    Hannah Aizenman\\
    Advisor: Dr. Michael Grossberg \\
    Committee: Dr. Robert Haralick, Dr. Lev Manovich, Dr. Huy Vo\\
    External Member: Dr. Marcus Hanwell
\end{frame}

\section{Introduction}

@@

\section{Prototype}
\begin{frame}{TEAM driven rearchitecture of Matplotlib}
    \begin{itemize}
        \begin{itemize}
            \item topologically complex heterogenous data 
            \item target display spaces
        \end{itemize}
        \pause
        \item structure preserving maps from data to visual
            \begin{itemize}
                \item data and graphics have equivalent continuity
                \item properties are equivariant under monoid actions
            \end{itemize}  
        \pause
        \item composable into complex visualizations
    \end{itemize}
\end{frame}

@@

\section{References}
\begin{frame}[allowframebreaks]{References}
\printbibliography
\end{frame}
\appendix 
\subsection{Appendix}

>>>>

\begin{frame}
	\titlepage
    Hannah Aizenman\\
    Advisor: Dr. Michael Grossberg \\
    Committee: Dr. Robert Haralick, Dr. Lev Manovich, Dr. Huy Vo\\
    External Member: Dr. Marcus Hanwell
\end{frame}
\section{Introduction}

@@

\section{Prototype}
\begin{frame}{TEAM driven rearchitecture of Matplotlib}
    \begin{itemize}
        \item complex visualizations
        \pause 
        \item structure preserving maps from data to visual
        \begin{itemize}
            \item data and graphics have equivalent continuity
            \item properties are equivariant under monoid actions
        \end{itemize}  
        \pause
        \item fiber bundles are an abstraction
        \begin{itemize}
            \item topologically complex heterogenous data 
            \item target display spaces
        \end{itemize}
    \end{itemize}
\end{frame}

@@

\section{References}
\begin{frame}[allowframebreaks]{References}
\printbibliography
\end{frame}
\appendix

\subsection{Appendix}

\begin{frame}{Mathematical Models of Vizualization}
    \begin{description}
        \item[algebraic process] data and viz transforms are symmetric \cite{kindlmannAlgebraicProcessVisualization2014}
        \begin{equation*}
            \begin{tikzcd}[ampersand replacement=\&]
                \dtotal \arrow[d, "m"'] \arrow[r, "\nu"] \& \vtotal \arrow[r, "Q"]  \& \gtotal \arrow[d, "\varphi(m)"] \\
                \dtotal \arrow[r, "\nu"']                     \& \vtotal \arrow[r, "Q"'] \& \gtotal                  
            \end{tikzcd}
        \end{equation*}

        \item[language] APT and GoG: syntax, semantics, and grammar\\ 
        \cite{mackinlayAutomatingDesignGraphical1986,mackinlayAUTOMATICDESIGNGRAPHICAL1987,wilkinsonGrammarGraphics2005}
        \item[functional dependencies] relationship between components \cite{sugibuchiFramwork2009}
        \item[category theory]  \textit{understanding} = \textit{read} $\circ$ \textit{render} \cite{vickersUnderstandingVisualizationFormal2013}    
    \end{description}
\end{frame}

\begin{frame}{Fiber is all possible values a variable can be \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}}
    Given a space of all possible values \ftotal\
    \begin{equation*}
        \label{eq:data_types}
        \begin{tikzcd}[ampersand replacement=\&]
            \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] \& \ftotal \arrow[d, "\pi"] \\
            \fnames \arrow[r, "\fsection"']                  \& \ftypes       
        \end{tikzcd}
    \end{equation*}
    a fiber component is the restricted space $\ftotal_{\fsection(\fname)}$. 
    \begin{equation*}
        \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
    \end{equation*}
    \begin{description}
        \item[\ftypes] data types of the variables in the dataset 
        \item[\ftotal] disjoint union of all values of type $\ftype \in \ftypes$ 
        \item[\fnames] variable names, $\fname \in \fnames$
        \item[\fttype] \ftotal\ restricted to the data type of a named variable   
    \end{description}
\end{frame}

\begin{frame}{Monoid actions}
    A monoid \monoid\ is a set with
    \begin{description}
        \item[associative binary operator] $\ast:\monoid \times \monoid\rightarrow \monoid$
        \item[identity element] $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$. 
    \end{description}
    \begin{block}{left monoid action}
    A set \dfiber\ with an action\cite{nlab:action} $\bullet: \monoid\times \dfiber \rightarrow \dfiber$ with the properties:
        \begin{align*}
            \textbf{associativity}\;& \text{for all } f,g \in \monoid \text{ and } x\in \dfiber,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
            \textbf{identity}\;& \text{for all } x\in \dfiber, e\in \monoid,\,  e\bullet x = x 
        \end{align*}
    \end{block}
\end{frame}

\begin{frame}{Keeping track of sections with sheafs}
    Restriction maps of a sheaf describe how local $\iota^*\dsection$ can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}
    \begin{equation*}
        \label{eq:sheaf}
        \begin{tikzcd}[ampersand replacement=\&]
            \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             \& \dtotal \arrow[d, "\pi"']                  \\
            U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] \& \dbase \arrow[u, "\dsection"', bend right]
        \end{tikzcd}
    \end{equation*}
    The inclusion map $\iota: U \rightarrow \dbase$ pulls \dtotal\ over $U$ such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$.
\end{frame}

\begin{frame}{Rendering: Define a Pixel}
    \begin{columns}
        \column{0.5\textwidth}
        Given a pixel
        \begin{equation*}
        p=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]
        \end{equation*}
        the inverse map of the bounding box 
        \begin{equation*}
        \gbase_{p} ={\gsection_{xy}}^{-1}(p)
        \end{equation*}
        is a region $\gbase_p \subset \gbase$ such that 
        \begin{align}
            \scriptstyle r_p &= \scriptstyle \iint\limits_{S_p} \rho_r(s)ds^{2}\\
            \scriptstyle g_p &= \scriptstyle \iint\limits_{S_p} \rho_g(s)ds^{2}\\
            \scriptstyle  b_p &= \scriptstyle \iint\limits_{S_p} \rho_b(s)ds^{2}
        \end{align}
        yields the color of the pixel. 
        \column{0.5\textwidth}
        \begin{figure}[H]
            \includegraphics[width=\textwidth]{figures/math/render.png}
        \end{figure}
    \end{columns}
\end{frame}

\begin{frame}{$\mathcal{A}:\mathcal{E}\rightarrow\mathcal{E}$}
    The topological artist is a sheaf map
    \begin{equation*}
    \vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)
    \end{equation*}
    that carries homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ 
    \cite{cegarraCohomologyMonoidsOperators2019} 
    \begin{equation*}
    \vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
    \end{equation*}
\end{frame}

\begin{frame}{Visual Channel Encoders}
    We define the visual transformers \vchannel\ on components of the data bundle $\dsection_{i}$
    \begin{equation*}
        \label{eq:nu_expanded}
        \{\vchannel_{0}, \ldots, \vchannel_{n}\}: \{\dsection_{0}, \ldots, \dsection_{n}\} \mapsto \{\vsection_{0}, \ldots, \vsection_{n}\}
    \end{equation*}
    as the set of equivariant maps with the constraint 
    \begin{equation*}
        \vchannel_i(m_{\delement}(\dtotal_i)) = \varphi(m_{\delement})(\vchannel_i(\dtotal_i))
    \end{equation*} 
    where $\varphi:\monoid\rightarrow \monoid^{\prime}$ carries a homomorphism of monoid actions. 
\end{frame}

\begin{frame}{\vfiber\ Components}    
\begin{table}[H]
    \renewcommand{\arraystretch}{2}
    \begin{tabulary}{\textwidth}{|l|L|l|}\hline
     $\bm{\vchannel_{i}}$                      & $\bm{\vsection_{i}}$                                                            & $\bm{codomain(\vchannel_{i}) \subset \vfiber_{i}}$  \\ \hline                                              
    position                    & x, y, z, theta, r                                                          & $\mathbb{R}$   \\ \hline
    size                        & linewidth, markersize                                            & $\mathbb{R}^{+}$   \\ \hline
    shape                       & markerstyle                                                      & $\{f_{0}, \ldots, f_{n}\}$ \\ \hline
    color                       & color, facecolor, markerfacecolor, edgecolor  & $\mathbb{R}^{4}$ \\ \hline
    \multirow{2}{*}{texture}    & hatch                                                            & $\mathbb{N}^{10}$\\\cline{2-3}
                                & linestyle                                                        & $(\mathbb{R}, \mathbb{R^+}^{n, n\%2=0})$ \\ \hline              
    \end{tabulary}
    \label{tab:mpl_visual_variable_fiber}
\end{table}
\end{frame}

\begin{frame}{Glyph}
    The glyph is the graphic generated by $\vmark(\gbase_{\dbasepathpoint})$ where the path connected components $\dbasepath \subset \dbase$ are defined 
    \begin{equation*}
    \dbasepath = \{\dbasepathpoint \in \dbase \text{ s. t. } \exists \gamma \text{ s.t. } \gamma(0)=\dbasepoint \text{ and }\gamma(1)=\dbasepathpoint\}
    \end{equation*}
    such that the  path $\gamma$ from \dbasepoint\ to \dbasepathpoint\ is a continuous function from the interval [0,1] and $\gbase_{\dbasepathpoint}$ is the region
    \begin{equation*}
        \begin{tikzcd}[ampersand replacement=\&]
            \gtotal \arrow[r, shift left] \& \gbase_\dbasepathpoint \arrow[rr, "\vindex(\gbasepoint)", shift left] \arrow[l, "\gsection(\gbase_\dbasepathpoint)"] \& \& \dbasepath_{\dbasepoint} \arrow[ll, "\vindex^{-1}(\dbasepath)"]
            \end{tikzcd}
        \label{eq:mark}
    \end{equation*}
    such that the glyph is differentiable, in keeping with Ziemkiewicz and Kosara's description of a glyph\cite{ziemkiewiczEmbeddingInformationVisualization2009}.
\end{frame}

\begin{frame}{Artist Equivalance class}
    When artists share a base space 
    \begin{equation*}
        \dbase_2 \hookrightarrow \dbase_1
    \end{equation*}
    a composition operator can be defined such that the the artists can be considered to be acting on different components of the same section. 
\end{frame}

\begin{frame}[fragile]{Complex \vchannel}
    \begin{minted}{python}
    class Categorical:
        def __init__(self, mapping):
            # check that the conversion is to valid colors
            assert(mcolors.is_color_like(color) for color in mapping.values())
            self._mapping = mapping

        def __call__(self, value):
            # convert value to a color
            return [mcolors.to_rgba(self._mapping[v]) for v in values]
    \end{minted}
    That we can test for action equivariance
    \begin{minted}{python}
    def test_nominal(values, encoder):
        m1 = list(zip(values, encoder(values)))
        random.shuffle(values)
        m2 = list(zip(values, encoder(values)))
        assert sorted(m1) == sorted(m2)
    \end{minted}
\end{frame}

\begin{frame}[fragile]{Artist}
    \begin{minted}{python}
        class ArtistClass(matplotlib.artist.Artist):
            def __init__(self, data, transforms, *args, **kwargs):
                # properties that are specific to the graphic
                self.data = data 
                self.transforms = transforms
                super().__init__(*args, **kwargs)

            def assemble(self, **args):
                # set the properties of the graphic

            def draw(self, renderer):
                # returns K, indexed on fiber then key 
                view = self.data.view(self.axes) 
                # visual channel encoding applied fiberwise 
                visual = {p: t['encoder'](view[t['name']])
                          for p, t in self.transforms.items()}
                self.assemble(**visual)
                # pass configurations off to the renderer
                super().draw(renderer)
        \end{minted}
\end{frame}

\begin{frame}[fragile]{Artists: Scatter \& Line}
\begin{minted}{python}
class Point(mcollections.Collection):
    def assemble(self, x, y, s, facecolors='C0' ):
        # construct geometries of the circle glyphs in visual coordinates
        self._paths = [mpath.Path.circle(center=(xi,yi), radius=si) 
                    for (xi, yi, si) in zip(x, y, s)] 
        # set attributes of glyphs, these are vectorized 
        # circles and facecolors are lists of the same size
        self.set_facecolors(facecolors)

class Line(mcollections.LineCollection):
    def assemble(self, x, y, color='C0'):
        #assemble line marks as set of segments 
        segments = [np.vstack((vx, vy)).T for vx, vy in zip(x, y)]
        self.set_segments(segments)
        self.set_color(color)
\end{minted}
\end{frame}

\begin{frame}[fragile]{View}
\begin{minted}{python}
def view(self, axes):
    table = defaultdict(list)
    for k in self.keys:
        table['index'].append(k)
        for (name, value) in zip(self.FB.fiber.keys(), self.tau(k)[1]):
            table[name].append(value)
    return table
\end{minted}
\begin{description}
    \item [\mintinline{python}{VertexSimplex}] (name, value), value is scaler
    \item [\mintinline{python}{EdgeSimplex}]  (name, value), value is [x0, ..., xn]
\end{description}
\end{frame}

\begin{frame}[fragile]{Fiber Bundle}
    \begin{minted}{python}
    @dataclass
    class FiberBundle:
    """
    Attributes
    ----------
    K: {'tables': []}
    F: {variable name: type}
    """
        K: dict 
        F: dict
    \end{minted}
\end{frame}

\begin{frame}[fragile]{GraphLine Data Model}
\begin{minted}{python}
class GraphLine:
    def __init__(self, FB, edge_table, vertex_table, num_samples=1000,
                        connect=False):
        #set args as attributes and generate distance
        if connect: # test connectivity if edges are continuous
            assert edge_table.keys() == self.FB.F.keys()
            assert is_continuous(vertex_table)

    def tau(self, k):
        # evaluates functions defined in edge table
        return(k, (self.edges[c][k](self.distances) 
                        for c in self.FB.F.keys()))

    def view(self, axes):
        # walk the edge_vertex table to return the edge function
        table = defaultdict(list)
        for (i, (start, end)) in sorted(zip(self.ids, self.vertices), 
                                            key=lambda v:v[1][0]):
            table['index'].append(i)
            # same as view for line, returns nested list
            for (name, value) in zip(self.FB.F.keys(), self.tau(i)[1]):
                table[name].append(value)
        return table
\end{minted}
\end{frame}
